What happens to the inhibitory control functions of the right inferior frontal cortex when this area is dominant for language?

Version 1

Abstract

Abstract

A low number of individuals show an atypical brain control of language functions that differs from the typical lateralization in the left cerebral hemisphere. In these cases, the neural distribution of other cognitive functions is not fully understood. Although there is a bias towards a mirrored brain organization consistent with the Causal hypothesis, some individuals are found to be exceptions to this rule. However, no study has focused on what happens to the homologous language areas in the right frontal inferior cortex. Using an fMRI-adapted stop-signal task in a healthy non right-handed sample (50 typically lateralized and 36 atypically lateralized for language production), our results show that atypical lateralization is associated with a mirrored brain organization of the inhibitory control network in the left hemisphere: inferior frontal cortex, presupplementary motor area, and subthalamic nucleus. However, the individual analyses revealed a large number of cases with a noteworthy overlap in the inferior frontal gyrus, which shared both inhibitory and language functions. Further analyses showed that atypical lateralization was associated with stronger functional and structural interhemispheric connectivity.

Importantly, we did not find task performance differences as a function of lateralization, but there was an association between atypical dominance in the inferior frontal cortex and higher scores on schizotypy and autistic spectrum traits, as well as worse performance on a reading accuracy test. Together, these results partially support the Causal hypothesis of hemispheric specialization and provide further evidence of the link between atypical hemispheric lateralization and increased interhemispheric transfer through the corpus callosum.

Introduction

One of the oldest findings in human neuroscience is that the brain control of language is lateralized to the left hemisphere1–3. However, not all individuals present a typical organization of this function. Some left-handers (22-24%) show an atypical lateralization of language – right or ambilateral4–6. Also, this atypical organization is observed in some neurodevelopmental disorders such as schizophrenia, dyslexia, or autistic spectrum7–9.

How these atypical individuals organize the rest of their lateralized functions has been an extensively researched question. Historically, two different hypotheses have been proposed to explain the hemispheric specialization of the brain. The Causal hypothesis10, 11 argues that rapid activities requiring the implementation of sequences of cognitive processes are better performed from a single, unilateral control. As a result, these activities are innately programmed to be lateralized, and, according to this model, some functions will be performed better if they are controlled by different hemispheres. The Statistical hypothesis12, on the other hand, postulates that each cognitive function lateralizes independently from the others, making the probability of finding an atypical hemispheric dominance merely a statistical phenomenon. The evidence supporting each theory is mixed. Some studies have obtained data favoring the Statistical hypothesis when analyzing language, visuospatial attention, and face processing13–15. Most investigations, however, have demonstrated a mirrored organization of praxis16, 17, visuospatial attention14, 17, 18, face recognition14, 17, 19, and emotional prosody17 in most individuals with atypical lateralization of language. Nonetheless, it should be noted that this reversed pattern is not observed in all atypical participants, and all the studies found specific cases in favor of both theories17.

All these studies have employed a production task to investigate language lateralization in the inferior frontal cortex (IFC), classifying left-handed participants into typical and atypical groups. The rest of the lateralized cognitive functions have been examined using several tasks, but they all mainly depended on posterior parts of the brain. This may have limited the results, given that they were not dealing with directly homotopic functions, which would have been the most sensitive scenario to the proposed advantages of hemispheric specialization20, 21. To directly address the language production function, which is typically dependent on the left IFC, we would have to use a cognitive task that is typically dependent on the homotopic right IFC. According to previous literature, this function would be inhibitory control. Extensive results support the notion that stopping an ongoing response involves a rightward network that includes the IFC, presupplementary motor area (preSMA), and subthalamic nucleus (STN)22–25. Thus, inhibitory control and language production together are the ideal candidates to investigate the Causal vs. Statistical disjunctive. But surprisingly, no investigation to date has analyzed this.

In this study, we explored the relationship between the lateralization of language production and inhibitory control. To do so, we divided non-right- handed participants into typically lateralized and atypically lateralized for language using the verb generation task. These two groups then completed the stop-signal task. We hypothesized that the atypical group would show a mirrored brain organization during the stop-signal task. Additionally, we investigated if a relation existed between the lateralization of both functions and inter-hemispheric structural-functional connectivity26, as well as with behavioral markers of certain clinical conditions that have been related with atypical lateralization7–9.

Method

Participants

Eighty-six participants were included in the present study. They were selected following a functional magnetic resonance imaging (fMRI) language lateralization assessment via a verb generation task. Hence, 50 were typically lateralized for language – left-dominant – (mean ± SD age = 22.4 ± 3.6 years; 24 male, 26 female) and 36 were atypically lateralized (mean ± SD age = 23.4 ± 4 years; 16 male, 20 female).

All the participants were non-right-handed, according to the Edinburgh Handedness Inventory27, 28. There were no significant between-group differences in age (t84 = 1.28; P = 0.2), sex (χ2 =0.11; P = 0.74), or general intelligence by WAIS-IV matrix reasoning subtest29 (t84 = 1.8; P = 0.08). The participants had no history of any neurological or psychiatric disorders or head injury with loss of consciousness. Written informed consent was obtained from all participants following a protocol approved by Universitat Jaume I. All methods were carried out in accordance with approved guidelines and regulations.

Experimental Design

Participants were recruited via multiple advertisements across Castellón and Valencia universities (bulletin boards, mass emailing, etc.) asking for the collaboration of left-handers in an fMRI brain study. Persons older than 36 years or younger than 16 years were discarded. Valid participants were cited for a first fMRI session, in which they completed the verb generation task and the resting- state acquisition. During this session, we used the BrainWave software (GE HealthCare Technologies Inc.) to visualize real-time data and roughly categorize participants as potentially typical or potentially atypical. 43 participants were found potentially atypical for language lateralization, and were invited to a second fMRI session, along with 43 potentially typical participants. During this second session, participants completed the stop-signal task (scanner) and a reading skill test (out of scanner). Also, data regarding schizotypy personality and autistic spectrum was gathered via self-questionnaires. Note that classification of participants as potentially typical or potentially atypical was used for screening purposes, and it did not completely match the final assessment.

Verb generation Task

Expressive language function was measured via fMRI during a computerized verb generation task30, 31 that consists of a 2-block design paradigm with intercalating activation and control blocks. In the activation blocks, participants are asked to overtly say the first verb that comes to mind when visually presented with a concrete noun. In the control blocks, they have to read visually-presented letter pairs aloud. This task lasted for 6 minutes, with a block duration of 30 seconds (6 activation blocks and 6 control blocks), a stimulus duration of 1500 ms, and a blank inter-stimulus interval of 1500 ms. Before entering the scanner, participants practiced with a different version of the task for 1 minute. Stimuli were presented using MRI-compatible goggles (VisuaStim Digital, Resonance Technology Inc.), and responses were recorded with a noise- cancelling microphone (FOMRI III+, Optoacoustics Ltd.) to verify that each participant was engaged correctly in the task.

Stop-signal Task

Response inhibition was measured via fMRI during a computerized stop-signal task adapted from another study23. The task consists of an event-related design paradigm with ‘Go’ and ‘Stop’ trials. In the ‘Go’ trials, participants are asked to manually answer by pressing a button if the visually presented string is a word (has a meaning) or a pseudoword (mimics typical word structure but has no meaning). Words require an index button press, whereas pseudowords require a thumb button press. However, on the ‘Stop’ trials, the string is followed by a ‘beep’ noise that signals to the participants to inhibit their response and refrain from pressing any button. The instructions emphasized the importance of going correctly and stopping correctly. However, participants were asked to respond as quickly and accurately as possible and avoid withholding their responses in anticipation of a possible ‘beep’.

This task took 13 minutes and 28 seconds, and it was divided into two runs that lasted 6 minutes and 14 seconds each, separated by a 1-minute rest. Each run consisted of 135 trials, of which 32 (23.7%) were ‘Stop’ trials. All the trials randomly presented words or pseudowords. The trial structure consisted of a fixation crosshair (500 ms), a word or pseudoword (1000 ms, during which the ‘beep’ may or may not be presented at some point), and a blank inter-stimulus interval (ranging from 500 to 4000 ms, sampled from an exponential distribution truncated at 4000 ms, with mean of 1000 ms). The Stop Signal Delay or SSD (the amount of time after the onset of the word or pseudoword when the ‘beep’ is presented during ‘Stop’ trials) changed dynamically during the task after each ‘Stop’ trial, depending on whether inhibition was successful (+25 ms) or unsuccessful (−25 ms), with a minimum of 100 ms and a maximum of 800 ms.

The lower the SSD, the easier it is to inhibit the response, and vice versa. Hence, a dynamic SSD normalizes the task’s difficulty across all participants based on their performance, aiming at a 50% successful inhibition rate. The SSD used for the first ‘Stop’ trial was estimated for each participant based on their practice session before entering the scanner. This practice session lasted 6 minutes and used a different set of words and pseudowords. Stimuli were presented using MRI-compatible goggles and headset (VisuaStim Digital, Resonance Technology Inc.), and responses were recorded via an MRI-compatible response-grip (ResponseGrips, NordicNeuroLab). The ‘beep’ volume was kept at a comfortable level and was constant across participants.

Resting-state

Functional connectivity was measured via fMRI during a resting-state session32. In this paradigm, participants were presented with a fixation crosshair and instructed to just lie in the scanner with their eyes open and try not to sleep or think about anything in particular. This session lasted for 7 minutes. Seven participants did not complete this session due to time constraints, and they were subsequently removed from the functional connectivity analyses.

Behavioral measures

Individual inhibition speed was estimated by calculating the Stop-Signal Reaction Time (SSRT)33. SSRT was computed as the difference between the median reaction time (RT) on correct ‘go’ trials and the mean SSD on the stop- signal task. It should be noted that technical problems involving the MRI- compatible response-grip invalidated the SSD data of four participants, and so their SSRT was estimated using the SSD data from the practice session instead.

Reading skill was evaluated using the word reading subtest of the PROLEC-SE- R34, a battery that assesses reading processes in Spanish. Participants had to read four lists of words consisting, respectively, of short familiar words, long familiar words, short unfamiliar words, and long unfamiliar words. Responses were recorded with a microphone, and their accuracy and speed were subsequently measured by computing length (performance during short words – performance during long words) and familiarity (performance during familiar words – performance during unfamiliar words) indicators.

Schizotypal traits were explored with the SPQ35, a preclinical self-report questionnaire modeled after DSM-III criteria36. This questionnaire evaluates schizotypy based on the three-dimensional model: disorganized traits, cognitive- perceptual traits, and interpersonal traits. For two participants, these data were not collected due to time constraints.

Autistic spectrum traits were explored with the AQ37, a preclinical self-report questionnaire modeled after DSM-IV criteria38. This questionnaire evaluates autistic spectrum based on the “triad” of impaired communication, impaired social skills, and a restricted and repetitive way of acting39, plus two sub-scales for imagination and attention to detail. For two participants, these data were not collected due to time constraints.

Image acquisition and processing

Images were acquired on a 3T General Electric Signa Architect magnetic resonance imaging (MRI) scanner. A 3D structural MRI was acquired for each subject using a T1-weighted magnetization-prepared rapid gradient-echo sequence (TR/TE = 8.5/3.3 ms; matrix = 512 × 512 × 384; voxel size = 0.47 × 0.47 × 0.5). For the fMRI, a gradient-echo T2*-weighted echo-planar imaging sequence was used in the acquisition of 150 functional volumes on the verb generation task (TR/TE = 2500/30 ms; matrix = 64 × 64 × 30; voxel size = 3.75 × 3.75 × 4), and a different sequence was used in the acquisition of 374 functional volumes during the stop-signal task and 210 functional volumes in the resting- state (TR/TE = 2000/30 ms; matrix = 64 × 64 × 27; voxel size = 3.75 × 3.75 × 4.5).

Task functional images were processed using the Statistical Parametric Mapping software package (SPM12; Wellcome Trust Centre for Neuroimaging, London, UK). Preprocessing followed the default pipeline and included: (a) alignment of each participant’s fMRI data to the AC-PC plane by using the anatomical image; head motion correction, where the functional images were realigned and resliced to fit the mean functional image; (c) co-registration of the anatomical image to the mean functional image; (d) re-segmentation of the transformed anatomical image using a symmetric tissue probability map; (e) spatial normalization of the functional images to the MNI (Montreal Neurological Institute, Montreal, Canada) space with 3 mm3 resolution; and (f) spatial smoothing (FWHM = 4 mm). The general linear model for the verb generation task was defined for each participant by contrasting activation > control blocks.

The general linear model for the stop-signal task was defined for each participant by contrasting correct ‘stop’ > correct ‘go’ trials. For both tasks, the BOLD (Blood-Oxygen-Level-Dependent) signal was estimated by convolving the task’s block/trial onsets with the canonical hemodynamic response function. Six motion realignment parameters extracted from head motion preprocessing were included as covariates of no interest, and a high-pass filter (128 s) was applied to the contrast images to eliminate low-frequency components. After that, mean BOLD values were extracted from the stop-signal contrast images for the main components of the inhibitory control network: pars opercularis, pars triangularis, preSMA, and STN. Pars opercularis and pars triangularis of the IFC were defined following the criteria of the Harvard-Oxford atlas40–43. For the PreSMA region, we used the SMA region, as defined by the same atlas, but including only voxels anterior to MNI Y = 044. STN was defined by a 10mm box centered at MNI coordinates 10, −15, −522, 45. Lastly, we computed whole-brain voxel-wise functional asymmetry maps from the stop-signal contrast images. To do so, stop-signal contrast images were flipped at midline, inverting the right and left hemispheres, and subsequently subtracted from the original unflipped contrast images46.

Resting-state functional images were processed using the Data Processing Assistant for the Resting-State toolbox (DPARSFA)47, which is based on SPM and the Data Processing & Analysis of Brain Imaging toolbox (DPABI)48. Preprocessing steps included: (a) slice-timing correction for interleaved acquisitions; (b) head motion correction (no participant had a head motion of more than 2 mm maximum displacement in any direction or 2° of any angular motion throughout the scan); (c) co-registration of the anatomical image with the mean functional image; (d) new segmentation to DARTEL; (e) removal of nuisance variance through linear regression: six parameters from the head motion correction, white matter signal, cerebrospinal fluid signal, and global mean signal; (f) spatial normalization to the MNI (3 mm3); (g) spatial smoothing (FWHM = 4 mm); (h) removal of the linear trend in the time series; (i) band-pass temporal filtering (0.01–0.1); (j) normalization to a symmetric template; (k) Voxel-Mirrored Homotopic Computation (VMHC)49, calculated as the Pearson correlation coefficient of every voxel with its hemispheric counterpart; and (l) normalization of all voxel-wise time courses to Fisher z values. Finally, mean VMHC values were extracted for the pars opercularis and pars triangularis regions, as defined by the Harvard-Oxford atlas40–43.

Structural images were processed via voxel-based morphometry (VBM) using the CAT12 toolbox50, which is based on SPM. Preprocessing steps followed the recommended pipeline and included: (a) segmentation into grey matter, white matter, and cerebrospinal fluid; (b) registration to the ICBM standard template; modulated normalization of grey matter and white matter segments to the MNI template; (d) spatial smoothing (FWHM = 6 mm); and (e) extraction of Regions of Interest (ROIs) values from native space. Three ROIs were delimited by the voxels mapped as genu, body and splenium of the corpus callosum according to the Mori atlas51. Total intracranial volume was also extracted for use as covariate of no interest.

Individual assessment of functional lateralization

Individual functional lateralization was assessed by calculating the Laterality Index (LI) on the unflipped contrast images30, 31. We used the bootstrap method implemented in the LI-toolbox52, based on SPM. LI is a proportion of the brain activation between the two hemispheres, thus giving us information about the direction and degree of hemispheric specialization during a particular function in a single individual. LI ranges from +1 (totally leftward function) to −1 (totally rightward function). For both the verb generation task and the stop-signal task, we explored the LI of the inferior frontal region roughly corresponding to the classic Broca’s area: pars opercularis and pars triangularis of the inferior frontal gyrus, according to the Harvard-Oxford atlas40–43. This region is critical for both language production3 and response inhibition24, depending on the hemisphere. During the verb generation task, language production was classified as typically lateralized (LI higher than +0.4) or atypically lateralized (LI lower than +0.4). We used +0.4 as a cut-off point (contrary to the more traditional +0.2), based on previous findings that emphasized the importance of lateralization strength when grouping individuals6,53.

Statistical analyses

We performed different analyses to investigate the hypothesis that the atypical group would show a mirrored brain organization during the stop-signal task.

First, a series of ROI analyses were employed to compare the functional asymmetry of the main components of the inhibitory control network (during the stop-signal task) between typically and atypically lateralized groups. Hence, four repeated-measures MANOVAs (one for the mean BOLD value inside each ROI: pars opercularis, pars triangularis, preSMA, and STN) were calculated, including Hemisphere (left/right) as within-subject factor and Group (typical/atypical) as between-subject factor. All p-values were FDR-corrected for four comparisons. This analysis was complemented with whole-brain functional asymmetry contrast maps from the stop-signal task that were compared in typically and atypically lateralized groups based on language production. To do so, we performed a whole-brain two-sample t-test (voxel-wise P < 0.001; FWE cluster-corrected at P < 0.05) via SPM12. Next, a Pearson’s correlation was carried out between the LI from the verb generation task and the LI from the stop-signal task used in the individual lateralization classification (P < 0.05, 1- tailed). Finally, we explored the functional overlap between both tasks in the subgroup of participants that presented an ambilateral organization of at least one function (n = 48). To do so, we overlayed two one-sample t-tests restricted to the IFC: one for the stop-signal task (voxel-wise P < 0.001; FWE cluster- corrected at P < 0.05), and one for the verb generation task (voxel-wise P < 0.05; uncorrected). Functional overlap was then expressed as the proportion of overlapped voxels for every function and hemisphere.

Using the LIs from the verb generation and stop-signal tasks, we tested if a linear relationship existed between the functional lateralization of an individual, and several behavioral tests and neuroanatomical measures. Partial Spearman’s correlations were computed between the LI of both tasks and: stop-signal task performance (RT, accuracy, and SSRT), SPQ derived-scores, AQ derived-scores, reading test performance (length and familiarity effects on speed and accuracy), functional connectivity measures (VMHC), and white matter volume of the callosal genu. Age and general intelligence were included as covariates in all analyses. Total intracranial volume was also included as covariate in the volumetric analysis.

Results

Inhibitory control shifts its lateralization according to language production

Inhibitory control components during the stop-signal task were defined by four Regions Of Interest (ROIs) and subsequently analyzed by computing MANOVAs, including Hemisphere as within-subject factor and Group (typical/atypical) as between-subject factor. Results showed that all the tested structures presented a statistically significant Hemisphere × Group interaction in their BOLD signal during the ‘correct stop > correct go’ condition (pars opercularis F1,84 = 25.59, P < 0.001; pars triangularis F1,84 = 19.95, P < 0.001; preSMA F1,84 = 7.63, P = 0.007; STN F1,84 = 14.24, P < 0.001; FDR-corrected for 4 comparisons; see Fig 1a). In the atypical group, all the regions seemed to have a shift towards leftward lateralization of brain function.

Hemispheric lateralization of inhibitory control according to language lateralization. (A) ROI analysis of the main components of the inhibitory control network. Graphs depict adjusted mean BOLD signal during successful inhibitions on the stop-signal task (‘stop > go’ contrast) for both hemispheres and both groups. All four structures showed significant Hemisphere × Group interactions. Error bars represent one standard error. pOper = pars opercularis, pTri = pars triangularis, preSMA = presupplementary motor area, and STN = subthalamic nucleus. (B) Voxel-wise whole-brain analysis of functional asymmetry. Significance maps (voxel-wise P < 0.001; FWE cluster-corrected at P < 0.05; color bar represents t value) are displayed in three-dimensional reconstructions plus coronal and transversal slices using MNI space. (C) Mean BOLD values of the significant regions found in the voxel-wise whole-brain analysis. Graphic depicts BOLD values for every region, hemisphere, and group. IFC = inferior frontal cortex, preSMA = presupplementary motor area, AngG = angular gyrus, MTG = middle temporal gyrus, Thal = thalamus, STN = subthalamic nucleus, Caud = caudate. (D) Functional overlap between language production and inhibitory control in the IFC of ambilateral participants. Conjunction maps for inhibition (voxel-wise P < 0.001; FWE cluster-corrected at P < 0.05) and language (voxel-wise P < 0.05; uncorrected) are displayed in coronal, sagittal and transversal slices using MNI space. VGT = verb generation task, SST = stop-signal task.

A voxel-wise whole-brain analysis of the functional asymmetry maps during the stop-signal task was performed to complement the ROI results. Two-sample t- test comparisons of the typical and atypical groups during the ‘correct stop > correct go’ condition showed a differential asymmetry pattern in all the ROI- corresponding areas, while also revealing differences in other cortical and subcortical regions, including the angular gyrus, middle temporal gyrus, and caudate (Fig 1b). Specifically, the typical group presented a rightward activity in these regions, whereas the atypical group exhibited a leftward shift (Fig 1c).

Finally, the Pearson’s correlation between the frontal LIs (pars opercularis + pars triangularis) from the verb generation task and the stop-signal task revealed a strong inverse correlation (r84 = −.58; one-tailed P < 0.001; R2 = .34) (Fig 2). Out of 50 typically lateralized for language, 33 (66%) presented a typical right lateralization for inhibitory control, and 17 (34%) showed an atypical (left –ambilateral) lateralization. Among the 36 individuals from the atypical group, only 8 (22.2%) presented a typical lateralization during the stop-signal task, whereas 28 (77.8%) showed an atypical lateralization. Importantly, no participant presented a leftward + leftward or rightward + rightward segregation pattern for language production and response inhibition. Strong segregation of both functions into different hemispheres was observed in 38 participants (44.2%). The remaining 48 individuals (55.8%) presented an ambilateral representation of one or both functions, with an important level of spatial overlap in their activations (see Fig 1d). In this ambilateral group, language production presented an overlap in 32.3% of its voxels in the left hemisphere, and in 58.6% of its voxels in the right hemisphere. Inhibitory control showed an overlap in 57.2% of its voxels in the left hemisphere, and in 22.8% of its voxels of the right hemisphere.

Correlation between the LIs of the verb generation and stop-signal tasks. r = −.583, one-tailed P < 0.001, R2 = .339. Negative values indicate rightward lateralization, whereas positive values indicate leftward lateralization. Segregated and ambilateral phenotypes are also depicted according to the background color. The green area corresponds to segregated individuals (both functions strongly lateralized), and the red area corresponds to integrated individuals (at least one function ambilaterally controlled). Numbers inside each quadrant denote the number of individuals (n) contained in it. Each individual datapoint is symbolized according to its functional organization: 〇 = typical segregation; ⬤ = reversed segregation; △ = ambilateral inhibition; ▽ = ambilateral language; ◇ = ambilateral language and inhibition.

Laterality Indexes correlate to interhemispheric connectivity and preclinical markers

Correlational analyses revealed certain links between the functional lateralization of both tasks and behavioral, neuroanatomical and connectivity measures (Table 1).

First, we studied how the structural and functional connectivity between both IFC behaved in relation to LI. Volume in the genu, body and splenium of the corpus callosum was inversely related with language LI (respectively: ρ83 = −.25, P = 0.02; ρ83 = −.26, P = 0.02; and ρ83 = −.25, P = 0.02). However, only the callosal genu was found related with inhibition LI (ρ83 = .25, P = 0.02). That is, the volume of the callosal genu increased as functional organization of the IFC became more atypical, extending this effect to the whole corpus callosum when considering exclusively language lateralization. On the same line, ROI interhemispheric functional connectivity analyses at rest revealed that VMHC of the pars triangularis also increased as a function of inhibition LI (ρ77 = .26, P = 0.02), but a similar relation failed to be found for language LI (ρ77 = −.12, P = 0.3). No statistically significant linear relationships were found when exploring the VHMC of the pars opercularis.

Behaviorally, performance (RT, SSRT and accuracy) during the scanner stop- signal task did not correlate with language and inhibition LIs. Regarding the reading test, accuracy for word length and word familiarity were found correlated with both LIs (length and language: ρ84 = −.24, P = 0.03; length and inhibition: ρ84 = .21, P = 0.04; familiarity and language: ρ84 = −.23, P = 0.03; familiarity and inhibition: ρ84 = .3, P = 0.01). That is, an atypical organization was related with a higher rate of errors when reading long words and unfamiliar words. No significant correlations were found when exploring reading speed.

SPQ score was found significantly related with inhibition LI (ρ82 = .25, P = 0.02), but it did not reach statistical significance when paired with language LI (ρ84 = −.19, P = 0.08). AQ, on the other hand, presented an association with language LI (ρ84 = −.25, P = 0.02) but not with inhibition LI (ρ84 = .08, P = 0.5).

Discussion

The present study employed the stop-signal task to investigate brain areas involved in the inhibitory control function in two different groups of individuals: typically and atypically lateralized for language production. As expected, the atypical participants showed, as a group, a mirrored brain organization of the inhibitory control function compared to the typical group. This leftward organization affected the entire inhibitory control network, including the IFC, preSMA, and STN. However, some participants manifested a clear overlap in the control of language and inhibitory functions. Our results also demonstrate that atypical organization of language production is associated with an increased white matter volume of the corpus callosum, and that atypical lateralization of inhibitory control is related with a higher interhemispheric functional coupling of the IFC. Behaviorally, atypical lateralization was not associated with performance on the task, but an association existed with worse reading accuracy, and more schizotypy and autistic traits.

Our first relevant result is that the group of individuals with atypical lateralization of language presented a mirrored brain organization during the stop-signal task compared to the typically lateralized group. Remarkably, the hemispheric asymmetry differences obtained affected the entire inhibitory control brain network25. The inferior frontal cortex, or IFC, associated with the initiation of the stop command and the main cortical area of this system, showed a clear lateralization effect, with the typical group almost exclusively using the right part and the atypical group mostly activating the left part. The presupplementary motor area, or pre-SMA, related to the implementation of the stop command, was more bilaterally involved in both groups, although atypicals activated the left part more and typicals activated the right part more. Importantly, the hemispheric reversal of organization also affected the subcortical structures participating in this network, namely the STN and thalamus. These structures displayed a lateralization pattern similar to the preSMA, with a more pronounced bilateral engagement but higher involvement of the right part in typicals and of the left part in atypicals. Thus, this is the first demonstration that an entire cerebral network controlling a cognitive function shows a completely flipped organization in an atypical population for language, which supports the Causal hypothesis of lateralization10, 11. Moreover, this result reveals that hemispheric lateralization goes beyond cerebral cortices, and even interhemispheric connectivity, through the corpus callosum, given that it also involves frontal-subcortical circuits. Additionally, the specificity of this result to the areas of the proposed network is another element that supports the relevance of this system in inhibitory control.

It should be noted that this pattern of atypical brain specialization was observed in the absence of a relation between functional lateralization and performance differences during the ‘go’ or ‘stop’ conditions. Even though the ‘go’ condition involves a lexical decision requiring language processing, and the ‘stop’ condition consists of inhibiting a linguistic process, the functional organization of the IFC during the tasks did not affect the response speed (RT and SSRT) or the ‘go’ accuracy in either case. Thus, brain organization for language and inhibition, as initially defined in our study, and cognitive efficiency are not directly related, based on our data. This finding is consistent with previous reports showing that cognitive deficits appear only in specific cognitive domains17, 54.

Importantly, despite observing a complimentary relationship between language production and inhibitory control at the group level, not all the individuals presented a strong and segregated distribution of these two functions. The vast majority of hemispherically segregated individuals showed the typical segregation of language in the left hemisphere and inhibitory control in the right hemisphere, whereas only a low percentage displayed the opposite segregated organization. The rest, in agreement with similar reports on other functions19, lacked a clear hemispheric lateralization of language production, inhibitory control, or both. In all cases, this implied some level of overlapping or sharing of the same area of the inferior frontal cortex. Both the low ratio of atypical segregation and the high ratio of bilateral inhibitory control could be explained by the fact that typical right hemispheric functions (such as inhibition) seem to be more bilaterally represented than typical left functions (such as language)55. Some conclusions can be extracted from this: (1) Although infrequent, a strong atypical segregation of the two functions is possible; (2) Our data have not revealed any case with a segregation of both frontal cognitive functions in the same hemisphere; and (3) A large percentage of participants with a strong left lateralization of language presented a bilateral control of inhibition, showing that a typical dominance for language does not imply a strongly lateralized organization of other cognitive functions. The existence of exceptions to the group pattern of mirrored brain organization requires us to use caution when interpreting these data as supporting the causal hypothesis. Although the negative correlation between the lateralization indexes during the verb generation task and the stop-signal task is strong at the group level, some individual data support the statistical hypothesis12 instead. Given that we found no cases of both cognitive functions completely sharing a hemisphere (see Fig 2), we propose that causal-supporting vs. statistical-supporting results reflects two different pathways to cognitive control. On the one hand, the segregated pathway (in line with the Causal hypothesis) develops due to the cerebral bias towards lateralizing certain cognitive functions in a single hemisphere. On the other hand, the integrated pathway (in line with the Statistical hypothesis), appears when ontogenetic development towards strong lateralized control fails in some way.

The corpus callosum is the main cerebral structure in interhemispheric connectivity, with the genu area being responsible for the connectivity between the two inferior frontal gyri56. Phylogenetic data have demonstrated a negative association between the corpus callosum volume and hemispheric specialization, suggesting that the ontogenetic development of the brain is supported by a decrease in interhemispheric connectivity to potentiate and establish intrahemispheric connectivity and hemispheric specialization57. There is evidence of this mechanism in language development58 and in agenesis of the corpus callosum, which has been associated with lower lateralization of language and worse language performance59, 60. In line with this proposal, we have replicated the relation between callosal enlargement and atypical functional organization found in a previous study53. Moreover, in our data, this structural correlation was accompanied by an association between atypical lateralization of inhibitory control and strength of interhemispheric functional connectivity in the pars triangularis of the IFC.

As hypothesized, when considering the functional lateralization during both tasks, atypical organizations were found linearly related with higher scores on a schizotypy test (only for inhibitory control), higher scores on an autistic spectrum test (only for language production), and a higher rate of reading errors (both language production and inhibitory control). The first result is consistent with previous structural, functional, and behavioral data that have shown reduced language lateralization in schizophrenia patients8, 61. In the case of schizotypy, the evidence for reduced lateralization of cognitive functions is mixed and may respond to differences in the methodology employed62.

However, consistently with the present results, this condition has been previously associated to impaired behavioral and neural processing during the stop-signal task63, and to an abnormal frontal functional asymmetry64. The second result relates atypical language lateralization with the presence of autistic traits, a result consistent with data obtained in autism65. The third result shows that atypical organization was related with a higher rate of reading errors when considering word length and word familiarity. The length effect has been considered a pathognomonic symptom of reading disorders such as developmental dyslexia66 and pure alexia67. These disorders have also been characterized by a weak language lateralization7. Thus, presented results support the hypothesis that atypical hemispheric specialization is related with worse cognitive performance in the linguistic domain, and even preclinical traits of some neurodevelopmental disorders among healthy population.

Data availability

All data presented in this study are included in the following public dataset: https://doi.org/10.6084/m9.figshare.22047299.

eLife Assessment

This paper describes important advances from the results of studies on atypical language lateralisation and how this might impact language behaviour, as well as its associations with reading skills and the prevalence of schizotypy and autistic traits. Despite most of the conclusions of this paper being supported by evidence, several aspects including the characteristics of participants tested, hypotheses/predictions, and data analysis, are incomplete and/or need some clarification. The work has implications for evolutionary hypotheses and our understanding of the effects of brain damage on distinct cognitive functions.

Reviewer #1 (Public Review):

The authors' aim was to test to what extent atypical organization of language is associated with a mirrored brain organization of other cognitive functions. In particular, they focused on the inferior frontal gyri (IFG) by studying the inhibitory control network. This allowed them to directly test the support for the Causal hypothesis of hemispheric specialization, arguing for fast sequences of cognitive processes being better performed by a single hemisphere, versus the Statistical hypothesis of lateralization, postulating an independent lateralization of each cognitive function.

Previous studies on this topic did not focus on functions involving homotopic language regions. This limitation is bypassed in this study by assessing inhibition with a Stop-Signal Task which also engages the IFG in the contralateral site to the verb generation task. By studying a combination of structural and functional information, in addition to the activation contrasts, the authors are able to test whether atypical organization is accompanied by stronger interhemispheric connectivity. Although relying mainly on correlations and lacking important methodological information that may be critical to understand the reported effects, the results are quite straightforward. However the bilingual/monolingual status and gender of the participants is not reported which might affect the relationship between language and inhibitory control.

The conclusions of the paper are supported by the data. With their design, the authors observed that, as a group, individuals with atypical organization show a mirror organization of the whole inhibitory network to the contralateral site, supporting the Causal hypothesis at the group level. However, individual data support the Statistical hypothesis, since the segregation between language and inhibition was not observed in all individuals and a variety of configurations in bilateral and bilateral organisation of language and inhibition were also observed.

The results of this study have important implications for our understanding of the independence between different cognitive functions which is crucial when addressing brain damage and rehabilitation. This aspect also indirectly speaks to researchers interested in evolution and in bilingualism and its relation to cognitive control. These aspects are not discussed but incorporating them would broaden the interest of the paper beyond the current implications mentioned.

Reviewer #2 (Public Review):

Language skills are traditionally associated with a network of brain regions in the left hemisphere. In this intriguing study, Esteban Villar-Rodríguez and collaborators examined if atypical hemispheric lateralization for language determines the functional and structural organisation of the network for inhibitory control as well as its relationship with schizotypy and autistic spectrum traits. The results suggest that individuals who have atypical lateralisation of the language function have also an atypical (mirrored) lateralisation of the inhibitory control network, compared to the typical group (individuals with left-lateralised language function). Furthermore, the atypical organization of language production is associated with a greater white matter volume of the corpus callosum, and atypical lateralization of inhibitory control is related to a higher interhemispheric functional coupling of the IFC, suggesting a link between atypical functional lateralisation (language and inhibitory control) and structural and functional changes in the brain.

This study also provides interesting evidence on how atypical language lateralisation impacts some aspects of language behaviour (reading), i.e., atypical lateralization predicts worse reading accuracy. Furthermore, the results suggest an association between atypical lateralization and increased schizotypy and autistic traits.

The strength of this work is that it presents a collection of measurements on the same individuals (including task-related behavioural, functional and structural neuroimaging measures) to reveal if (and how) atypical language lateralisation might be associated with: (1) atypical neural organisation of other non-linguistic cognitive systems, (2) behavioural performance associated with language tasks, and finally (3) personality traits. As such the results presented in this manuscript have the potential to be informative for various disciplines. For instance, if clarifications/corrections are provided (see below), the results might provide some insight into the role of the right hemisphere for language processing in healthy individuals as well as patient populations with acquired linguistic impairment including stroke and dementia.

One important weakness of this manuscript is that several areas, including the characteristics of participants tested, and the hypotheses/predictions, are underspecified or incomplete. Furthermore, in some cases the types of analysis do not seem to be appropriate for addressing the questions of the present study and very little explanation for those choices is provided.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #2 (Recommendations For The Authors):

The evidence provided in this study reflects important discoveries on language lateralisation and most of the conclusions of this paper are supported by evidence. However, there are several areas regarding the characteristics of participants tested, hypotheses/predictions and the type of analysis, that need to be clarified and/or corrected.

1. There is a substantial disconnection between the introduction and the methods/results section.

One reason is because of lack of consistency. One example refers to the fact that, in the introduction, only IFC is mentioned. However, the analyses carried out to examine neural activity in different groups focused on IFC as well as other brain regions related to inhibitory control. However, these areas were not mentioned at all in the introduction. Second and related to the above, the rationale for conducting certain types of analyses is not specified. Some brain analyses focus on IFC only. Instead, other analyses focus on several areas.

Another weakness is that there is not sufficient detail regarding the hypotheses/predictions and the specific types of analyses chosen to test these hypotheses/predictions. For example, there is no mention of resting state fMRI data in the introduction, but later we discover that this type of data was collected and analyzed. Even a brief mention of the inclusion of resting state data in the introduction would be beneficial. Along the same lines, by reading the methods section we find out that VBM analyses were conducted. But it is unclear why. What was the purpose of this data analysis? This should be clarified briefly in the introduction and then in the methods section. It remains unclear why resting state results would be particularly informative for addressing the research question of this study. Task-related brain connectivity seems a more appropriate choice. Additionally, it is not explained what comparisons and outcomes would be informative/expected to distinguish between the two mentioned competing hypotheses. This should be made clear.

Another aspect that lacks clarity is the authors' predictions when investigating the relationship "between the lateralization of both functions and inter-hemispheric structural-functional connectivity, as well as with behavioural markers of certain clinical conditions that have been related with atypical lateralization". The hypotheses are completely omitted in this section.

Thank you for bringing this to our attention. We concur with Reviewer #2 that our introduction was somewhat lacking in detail and assumed too much prior knowledge on the part of the reader. This, together with a lack of a clear presentation of our tested hypotheses, made the introduction have a poor connection with both the results and discussion sections, which hindered the understanding of the paper.

As a result, we have made some additions to enhance the exposition of the following areas: (1) the causal and statistical hypotheses of lateralization (Lines 55-65); and (2) the hypotheses regarding subclinical markers of neurological disorders and the corpus callosum (Lines 90-104).

Furthermore, we have extensively revised the final paragraph of the introduction (Lines 105-121) to provide a clearer and more coherent linkage between the drivers presented during the introduction, our hypotheses, and the subsequent analyses.

1. It is important to provide more information on the language background of the participants. Were the participants in this study Catalan-Spanish bilinguals? If so, it is crucial for the authors to mention this.

Language background of the participants has been added to the corresponding section (Lines 138-145).

In fact, previous studies, including several publications from the authors themselves (Garbin et al., 2010; Rodríguez-Pujadas et al., 2013; Anderson et al., 2018), have shown that there are qualitative differences between bilinguals and monolinguals in the neural circuitry underlying executive control. Across all these studies, it was consistently reported that bilingual individuals, when engaged in non-linguistic inhibitory control tasks, recruited a broader network of left-brain regions associated with language control, including the left IFC, in comparison to monolingual individuals. If the participants in this study were indeed bilinguals, it raises concern if the aim of the study is to generalize the conclusions on lateralization effects beyond the bilingual population.

Rodríguez-Pujadas, A., Sanjuán, A., Ventura-Campos, N., Román, P., Martin, C., Barceló, F., … & Ávila, C. (2013). Bilinguals use language-control brain areas more than monolinguals to perform non-linguistic switching tasks. PLoS One, 8(9), e73028.

Garbin, G., Sanjuan, A., Forn, C., Bustamante, J. C., Rodríguez-Pujadas, A., Belloch, V., ... & Ávila, C. (2010). Bridging language and attention: Brain basis of the impact of bilingualism on cognitive control. NeuroImage, 53(4), 1272-1278.

Anderson, J. A., Chung-Fat-Yim, A., Bellana, B., Luk, G., & Bialystok, E. (2018). Language and cognitive control networks in bilinguals and monolinguals. Neuropsychologia, 117, 352-363.

Indeed, we have thoroughly reported that, when compared to monolinguals, bilinguals exhibit a significant implication of left brain regions during switching and inhibition tasks. So, this is a legitimate concern. Unfortunately, the society from which our participants were drawn is primarily bilingual, encompassing both active and passive bilinguals. The monolingual sample in those previous studies consisted of university students originating from predominantly monolingual regions of Spain. Given this context, it is unsurprising that the current study has a rather limited number of monolinguals (n=8), with only 2 displaying atypical language lateralization. Thus, we cannot provide a reliable answer to the role of bilingualism status in our data. Consequently, we have included a comment on this limitation on the discussion (Lines 504-512).

1. Regarding the methods section, I have the following specific queries. The first is about the control condition in the verb generation task. I find it puzzling that the 'task' and 'control' conditions differ in terms of the number of words uttered. Could the authors please provide further clarification on this?

Thank you for raising this question. Regarding the control condition, it is important to note that the design of this task drew inspiration from previously published verb generation tasks for fMRI (Benson et al., 1999; Fitzgerald et al., 1997) and PET (Petersen et al., 1988). In the fMRI tasks, a fixation cross served as the control condition, while the PET study used word repetition as the control. We acknowledged that a mere fixation cross might not adequately control for the movement and visual-related activations inherent in the verb generation task. Conversely, word repetition could potentially engage the default mode network due to the repetition of the same simple task, which might not be suitable for a control condition, and it could be overly linguistic because it involves a word. Consequently, we aimed to strike a balance by employing a control condition that consisted of reading letters. This approach allowed us to control for movement and vision factors without invoking semantics. Thus, after careful consideration, we ultimately opted on the reading of two letters to equate the response to the vocalization length of generating a verb.

Although we understand the concern of single vs. two vocalizations, it is worth emphasizing that this version of the verb generation task had undergone prior testing to assess its suitability for determining language lateralization in both healthy and clinical populations (Sanjuan et al., 2010). In fact, this task has been an integral component of our lab’s standard presurgical assessment protocol, which has been used for nearly two decades in individually evaluating language function in over 500 patients with central nervous system lesions.

Benson, R. R., Fitzgerald, D. B., Lesueur, L. L., Kennedy, D. N., Kwong, K. K., Buchbinder, B. R., Davis, T. L., Weisskoff, R. M., Talavage, T. M., Logan, W. J., Cosgrove, G. R., Belliveau, J. W., & Rosen, B. R. (1999). Language dominance determined by whole brain functional MRI in patients with brain lesions. Neurology, 4(52), 798–809.

Fitzgerald, D. B., Cosgrove, G. R., Ronner, S., Jiang, H., Buchbinder, B. R., Belliveau, J. W., Rosen, B. R., & Benson, R. R. (1997). Location of Language in the Cortex: A Comparison between Functional MR Imaging and Electrocortical Stimulation. AJNR Am J Neuroradiol, 18, 1529–1539.

Petersen, S. E., Fox, P. T., Posner, M. I., Mintun, M., & Raichle, M. E. (1988). Positron emission tomographic studies of the cortical anatomy of single-word processing. Nature, 331(18), 585–589.

Sanjuán, A., Bustamante, J. C., Forn, C., Ventura-Campos, N., Barrós-Loscertales, A., Martínez, J. C., Villanueva, V., & Ávila, C. (2010). Comparison of two fMRI tasks for the evaluation of the expressive language function. Neuroradiology, 52(5), 407–415. https://doi.org/10.1007/s00234-010-0667-8

Second, it is mentioned that some participants were excluded from different tasks due to technical issues or time constraints. It is important to ensure that all the results can be attributed to the exact same sample of participants across all tasks.

We absolutely agree that excluding participants can be problematic when presenting the results of multiple sets of analyses. Therefore, we repeated all analyses while excluding the 7 participants that lacked resting-state data. All results remained virtually identical, with a few minor exceptions:

1. Region-wise analysis of the stop-signal task: Hemisphere × Group effect in the preSMA region is significant (uncorrected P = 0.019), but it does not survive Bonferroni correction (corrected P = 0.076)

2. Voxel-wise analysis of the stop-signal task: The Thalamus + STN and Caudate clusters are significant at the voxel level, but do not survive the cluster-based FWE correction. They do survive FDR correction, though.

3. Correlation between SPQ score and LI of the stop-signal task: This correlation weakens just behind statistical significance, with a P value of 0.053.

4. Correlation between reading variables and LIs of both tasks: Severe drops in P values are evident between both LIs and reading length accuracy (P = .111 and .133), as well as between verb generation LI and reading familiarity accuracy (P = .111). However, the association between the stop-signal LI and the reading length time is now significant (r = −.229, P = .042).

According to this, we have included this statement in the methods section: (Lines 218-220).“It is important to highlight that the exclusion of these seven participants across all analyses does not notably impact the overall results.“

It is unclear how the authors have estimated the RTs results from the practice trials. This requires more explanation. Also, why was the median used for the Go Reaction Time instead of the mean, when calculating the individual SSRT?

We adapted the procedure used by Xue et al. (2008), implementing their approach to calculate SSRT. This has been elaborated further (Lines 227-230), together with the use of practice trials (Lines 233-236).

Xue, G., Aron, A.R., and Poldrack, R.A. (2008). Common Neural Substrates for Inhibition of Spoken and Manual Responses. Cerebral Cortex 18, 1923–1932. 10.1093/CERCOR/BHM220.

On a final note, information about the different types of pre-processing and data analysis is all reported in the same paragraph. I think using subsections would increase the intelligibility of the section.

Thank you for this suggestion. We have added subsections in both the ‘image processing’ and ‘statistical analyses’ sections.

1. Data analysis and Interpretation of the results. It is unclear how the mean BOLD signal was extracted to conduct ROI analysis (Marsbar?).

Thank you for ponting this out. Indeed, we were not very accurate in the description of this procedure. We extracted the first eigenvariate via the VOI function within SPM12. This has been included in Lines 291-293.

I feel uneasy about the way results are corrected for multiple comparisons. For instance, it is mentioned that in the ROI analysis, all p-values were FDR-corrected for four comparisons, but it is unclear why. The correct procedure for supporting conclusions about the effect of specific brain would be to have 'brain region' (n=4) as another within-subject factor. Furthermore, the one-tailed correlation is appropriate but only when testing for the possibility of a relationship in one direction and completely disregarding the possibility of a relationship in the other direction. However, this does not seem to be the case here (see Introduction), so a two-tailed correlation would be more appropriate.

We agree with Reviewer #2 that presenting this analysis as a single MANOVA that includes a ‘Region’ factor is a more accurate approach. Consequently, we have made the aforementioned correction in the methods section (Lines 357-364) and the results section (Lines 395-406). The LI-LI one-tailed correlation was also changed to a two-tailed correlation in the methods section (Line 383), the results section (Line 417), and Figure 2 (Line 886).

I am quite confused about using the term interhemispheric connectivity to refer to the volume of the genu, body and splenium of the corpus callosum. In fact, the volumes of genu, body and splenium of the corpus callosum do not reflect a measure of how strongly RH and LH IFC are connected to each other.

We agree that using the term ‘interhemispheric connectivity’ when referring to callosal volume may be somewhat misleading. We have replaced every instance of this terminology throughout the paper.

Furthermore, it is unclear why in a set of analyses (ROI and whole brain analyses) the authors focus on brain responses in different ROIs but instead, in connectivity measures the focus is only on IFC.

Our initial rationale was to focus on regions that are prominently involved in language, particularly the IFC, for examining inter-hemispheric connectivity at rest.

However, upon more careful consideration, it is true that the preSMA is also implicated in the language network (Labache et al., 2018), and certain studies have reported an impact of STN stimulation on specific language skills (for a review, see Vos et al., 2021). Consequently, we have incorporated these two regions into the resting-state analysis, along with subsequent correlations with LIs (Table 1 and Lines 118, 321-322 & 449-452).

Labache, L., Joliot, M., Saracco, J., Jobard, G., Hesling, I., Zago, L., Mellet, E., Petit, L., Crivello, F., Mazoyer, B., & Tzourio-Mazoyer, N. (2018). A SENtence Supramodal Areas AtlaS (SENSAAS) based on multiple task-induced activation mapping and graph analysis of intrinsic connectivity in 144 healthy right-handers. Brain Structure and Function 2018 224:2, 224(2), 859–882. https://doi.org/10.1007/S00429-018-1810-2

Vos, S. H., Kessels, R. P. C., Vinke, R. S., Esselink, R. A. J., & Piai, V. (2021). The Effect of Deep Brain Stimulation of the Subthalamic Nucleus on Language Function in Parkinson’s Disease: A Systematic Review. Journal of Speech, Language, and Hearing Research, 64(7), 2794–2810. https://doi.org/10.1044/2021_JSLHR-20-00515

Minor corrections/comments:

It is unclear why in figure caption 1, the conjunction maps are mentioned even if formal conjunction analysis was not conducted.

This poor choosing of words has been replaced to ‘overlapping maps’.

Line 382. VHMC should be VMHC.

Fixed. Thank you.

Line 334. This sentence and especially its relationship with the results is not clear at all. What do you mean by 'This finding is consistent with previous reports showing that cognitive deficits appear only in specific cognitive domains'?

This has been clarified (Lines 521-525).

Version 2

Abstract

Abstract

A low number of individuals show an atypical brain control of language functions that differs from the typical lateralization in the left cerebral hemisphere. In these cases, the neural distribution of other cognitive functions is not fully understood. Although there is a bias towards a mirrored brain organization consistent with the Causal hypothesis, some individuals are found to be exceptions to this rule. However, no study has focused on what happens to the homologous language areas in the right frontal inferior cortex. Using an fMRI-adapted stop-signal task in a healthy non right-handed sample (50 typically lateralized and 36 atypically lateralized for language production), our results show that atypical lateralization is associated with a mirrored brain organization of the inhibitory control network in the left hemisphere: inferior frontal cortex, presupplementary motor area, and subthalamic nucleus. However, the individual analyses revealed a large number of cases with a noteworthy overlap in the inferior frontal gyrus, which shared both inhibitory and language functions. Further analyses showed that atypical lateralization was associated with stronger functional interhemispheric connectivity and larger corpus callosum. Importantly, we did not find task performance differences as a function of lateralization, but there was an association between atypical dominance in the inferior frontal cortex and higher scores on schizotypy and autistic spectrum traits, as well as worse performance on a reading accuracy test. Together, these results partially support the Causal hypothesis of hemispheric specialization and provide further evidence of the link between atypical hemispheric lateralization and increased interhemispheric transfer through the corpus callosum.

Introduction

One of the oldest findings in human neuroscience is that the brain control of language is lateralized to the left hemisphere1–3. However, not all individuals present a typical organization of this function. Some left-handers (22-24%) show an atypical lateralization of language – right or ambilateral4–6. Also, this atypical organization is observed in some neurodevelopmental disorders such as schizophrenia, dyslexia, or autistic spectrum7–9.

How these atypical individuals organize the rest of their lateralized functions has been an extensively researched question. Historically, two different hypotheses have been proposed to explain the hemispheric specialization of the brain. The Causal hypothesis10,11 argues that rapid activities requiring the implementation of sequences of cognitive processes are better performed from a single, unilateral control. As a result, these activities are innately programmed to be lateralized, and, according to this model, some functions will be performed better if they are controlled by different hemispheres. This evolutionary pressure engenders a complementary relationship between these lateralized functions. For example, language production predominantly lateralizes to the left hemisphere, while visuospatial attention typically favors the right hemisphere. As per the Causal hypothesis, if language were to atypically lateralize to the right hemisphere, visuospatial attention would, in turn, shift to the left hemisphere. The Statistical hypothesis12, on the other hand, postulates that each cognitive function lateralizes independently from the others. In this view, the likelihood of encountering atypical hemispheric dominance is essentially a statistical phenomenon, with no inherent interrelationship between various lateralized functions. The evidence supporting each theory in left-handers is mixed. Some studies have obtained data favoring the Statistical hypothesis when analyzing language, visuospatial attention, and face processing13–15. Most investigations, however, have demonstrated a mirrored organization of praxis16,17, visuospatial attention14,17,18, face recognition14,17,19, and emotional prosody17 in most individuals with atypical lateralization of language. Nonetheless, it should be noted that this reversed pattern is not observed in all atypical participants, and all the studies found specific cases in favor of both theories17.

All these studies have employed a production task to investigate language lateralization in the inferior frontal cortex (IFC), classifying left-handed participants into typical and atypical groups. The rest of the lateralized cognitive functions have been examined using several tasks, but they all mainly depended on posterior parts of the brain. This may have limited the results, given that they were not dealing with directly homotopic functions, which would have been the most sensitive scenario to the proposed advantages of hemispheric specialization20,21. To directly address the language production function, which is typically dependent on the left IFC, we would have to use a cognitive task that is typically dependent on the homotopic right IFC. According to previous literature, this function would be inhibitory control. Extensive results support the notion that stopping an ongoing response involves a rightward network that includes the IFC, presupplementary motor area (preSMA), and subthalamic nucleus (STN)22–25. Thus, inhibitory control and language production together are the ideal candidates to investigate the Causal vs. Statistical disjunctive. But surprisingly, no investigation to date has analyzed this.

What causes language function to typically or atypically lateralize also remains unclear. Left-handedness appears to be, at best, a predisposing factor6. In clinical populations, atypical lateralization of language has been found to be a defining characteristic among patients of dyslexia7, schizophrenia8, and autism spectrum disorders9. However, aside from exhibiting poorer language performance26,27, no study to date has identified subclinical markers of these neurodevelopmental conditions among healthy individuals with atypical lateralization. Regarding the neural basis of language lateralization, the corpus callosum hypothesis28 — which suggests that typical hemispheric specialization relies on the proper development of this structure — has garnered attention in recent years due to several noteworthy findings: (1) regional functional lateralization is directly linked to the size reduction of its interconnecting corpus callosum segments29; and (2) atypical lateralization of language has been associated with a larger corpus callosum and increased interhemispheric transfer at rest in language-relevant regions30.

The primary aim of this study was to investigate the relationship between the lateralization of language production and inhibitory control. To achieve this, we divided non-right-handed participants into typically lateralized and atypically lateralized for language using the verb generation task. These two groups then completed the stop-signal task. In alignment with the predictions of the Causal hypothesis, we expected that: (1) the atypical group would exhibit a mirrored brain organization — particularly in the IFC, preSMA, and STN — during the stop-signal task; and (2) no individual would demonstrate lateralization of both language production and inhibitory control in the IFC of the same hemisphere. The secondary aim of this study was to explore certain potentially causal correlates of the lateralization of both functions. To test the presence of correlates supporting the corpus callosum hypothesis of lateralization, we analyzed the callosal volume and the interhemispheric functional connectivity at rest of the IFC, preSMA, and STN. To test the relation between atypical lateralization and several neurodevelopmental disorders (dyslexia, schizophrenia, and autism), we analyzed subclinical traits of these conditions among our healthy participants.

Method

Participants

Eighty-six participants were included in the present study. They were selected following a functional magnetic resonance imaging (fMRI) language lateralization assessment via a verb generation task. Hence, 50 were typically lateralized for language – left-dominant – (mean ± SD age = 22.4 ± 3.6 years; 24 male, 26 female) and 36 were atypically lateralized (mean ± SD age = 23.4 ± 4 years; 16 male, 20 female).

All the participants were non-right-handed, according to the Edinburgh Handedness Inventory31,32. There were no significant between-group differences in age (t84 = 1.28; P = 0.2), sex (χ2 =0.11; P = 0.74), or general intelligence by WAIS-IV matrix reasoning subtest33 (t84 = 1.8; P = 0.08). The participants had no history of any neurological or psychiatric disorders or head injury with loss of consciousness. Written informed consent was obtained from all participants following a protocol approved by Universitat Jaume I. All methods were carried out in accordance with approved guidelines and regulations.

Regarding their bilingual status, we found three categories: Spanish monolinguals (n = 6 typical and 2 atypical), Spanish-Catalan bilinguals (n = 27 typical and 23 atypical); and Spanish-Catalan passive bilinguals (n = 17 typical and 11 atypical). Passive bilingualism refers to the fact that some residents of this region (Valencian Community) understand Spanish and Catalan, but their frequency of use of one language is extremely low when compared to the other. No significant between-group differences were found in bilingual status when comparing the typical and atypical groups (χ2 =1.2; P = 0.55).

Experimental Design

Participants were recruited via multiple advertisements across Castellón and Valencia universities (bulletin boards, mass emailing, etc.) asking for the collaboration of left-handers in an fMRI brain study. Persons older than 36 years or younger than 16 years were discarded. Valid participants were cited for a first fMRI session, in which they completed the verb generation task and the resting-state acquisition. During this session, we used the BrainWave software (GE HealthCare Technologies Inc.) to visualize real-time data and roughly categorize participants as potentially typical or potentially atypical. 43 participants were found potentially atypical for language lateralization, and were invited to a second fMRI session, along with 43 potentially typical participants. During this second session, participants completed the stop-signal task (scanner) and a reading skill test (out of scanner). Also, data regarding schizotypy personality and autistic spectrum was gathered via self-questionnaires. Note that classification of participants as potentially typical or potentially atypical was used for screening purposes, and it did not completely match the final assessment.

Verb generation Task

Expressive language function was measured via fMRI during a computerized verb generation task34,35 that consists of a 2-block design paradigm with intercalating activation and control blocks. In the activation blocks, participants are asked to overtly say the first verb that comes to mind when visually presented with a concrete noun. In the control blocks, they have to read visually-presented letter pairs aloud. This task lasted for 6 minutes, with a block duration of 30 seconds (6 activation blocks and 6 control blocks), a stimulus duration of 1500 ms, and a blank inter-stimulus interval of 1500 ms. Before entering the scanner, participants practiced with a different version of the task for 1 minute. Stimuli were presented using MRI-compatible goggles (VisuaStim Digital, Resonance Technology Inc.), and responses were recorded with a noise-cancelling microphone (FOMRI III+, Optoacoustics Ltd.) to verify that each participant was engaged correctly in the task.

Stop-signal Task

Response inhibition was measured via fMRI during a computerized stop-signal task adapted from another study23. The task consists of an event-related design paradigm with ‘Go’ and ‘Stop’ trials. In the ‘Go’ trials, participants are asked to manually answer by pressing a button if the visually presented string is a word (has a meaning) or a pseudoword (mimics typical word structure but has no meaning). Words require an index button press, whereas pseudowords require a thumb button press. However, on the ‘Stop’ trials, the string is followed by a ‘beep’ noise that signals to the participants to inhibit their response and refrain from pressing any button. The instructions emphasized the importance of going correctly and stopping correctly. However, participants were asked to respond as quickly and accurately as possible and avoid withholding their responses in anticipation of a possible ‘beep’.

This task took 13 minutes and 28 seconds, and it was divided into two runs that lasted 6 minutes and 14 seconds each, separated by a 1-minute rest. Each run consisted of 135 trials, of which 32 (23.7%) were ‘Stop’ trials. All the trials randomly presented words or pseudowords. The trial structure consisted of a fixation crosshair (500 ms), a word or pseudoword (1000 ms, during which the ‘beep’ may or may not be presented at some point), and a blank inter-stimulus interval (ranging from 500 to 4000 ms, sampled from an exponential distribution truncated at 4000 ms, with mean of 1000 ms). The Stop Signal Delay or SSD (the amount of time after the onset of the word or pseudoword when the ‘beep’ is presented during ‘Stop’ trials) changed dynamically during the task after each ‘Stop’ trial, depending on whether inhibition was successful (+25 ms) or unsuccessful (−25 ms), with a minimum of 100 ms and a maximum of 800 ms. The lower the SSD, the easier it is to inhibit the response, and vice versa. Hence, a dynamic SSD normalizes the task’s difficulty across all participants based on their performance, aiming at a 50% successful inhibition rate. The SSD used for the first ‘Stop’ trial was estimated for each participant based on their practice session before entering the scanner. This practice session lasted 6 minutes and used a different set of words and pseudowords. Stimuli were presented using MRI-compatible goggles and headset (VisuaStim Digital, Resonance Technology Inc.), and responses were recorded via an MRI-compatible response-grip (ResponseGrips, NordicNeuroLab). The ‘beep’ volume was kept at a comfortable level and was constant across participants.

Resting-state

Functional connectivity was measured via fMRI during a resting-state session36. In this paradigm, participants were presented with a fixation crosshair and instructed to just lie in the scanner with their eyes open and try not to sleep or think about anything in particular. This session lasted for 7 minutes. Seven participants did not complete this session due to time constraints, and they were subsequently removed from the functional connectivity analyses. It is important to highlight that the exclusion of these seven participants across all analyses does not notably impact the overall results.

Behavioral measures

Individual inhibition speed was estimated by calculating the Stop-Signal Reaction Time (SSRT)37. SSRT was computed as the difference between the median reaction time (RT) on correct ‘go’ trials and the mean SSD on the stop-signal task. We chose to employ the median RT instead of the mean RT, in accordance with the approach used in the study from which we adapted our stop-signal task23. This choice was made for two specific reasons: (1) when working with RT, using the median helps reduce the impact of outlier responses during the task; and (2) using the median RT allowed our sample’s SSRT to align more closely with a normal distribution. It should be noted that technical problems involving the MRI-compatible response-grip invalidated the SSD and RT data of four participants, and so their SSRT was estimated using the data from the practice session instead. This practice session, which lasted for 6 minutes, took place outside of the scanner in a different room, but using an identical set of response-grip. So, for these four participants, the resulting SSD and RT after those 6 minutes of practice was used for their SSRT calculation.

Reading skill was evaluated using the word reading subtest of the PROLEC-SE-R38, a battery that assesses reading processes in Spanish. Participants had to read four lists of words consisting, respectively, of short familiar words, long familiar words, short unfamiliar words, and long unfamiliar words. Responses were recorded with a microphone, and their accuracy and speed were subsequently measured by computing length (performance during short words – performance during long words) and familiarity (performance during familiar words – performance during unfamiliar words) indicators.

Schizotypal traits were explored with the SPQ39, a preclinical self-report questionnaire modeled after DSM-III criteria40. This questionnaire evaluates schizotypy based on the three-dimensional model: disorganized traits, cognitive-perceptual traits, and interpersonal traits. For two participants, these data were not collected due to time constraints.

Autistic spectrum traits were explored with the AQ41, a preclinical self-report questionnaire modeled after DSM-IV criteria42. This questionnaire evaluates autistic spectrum based on the “triad” of impaired communication, impaired social skills, and a restricted and repetitive way of acting43, plus two sub-scales for imagination and attention to detail. For two participants, these data were not collected due to time constraints.

Image acquisition

Images were acquired on a 3T General Electric Signa Architect magnetic resonance imaging (MRI) scanner using a 32-channel head coil. All slices were acquired in the sagittal plane. A 3D structural MRI was acquired for each subject using a T1-weighted magnetization-prepared rapid gradient-echo sequence (TR/TE = 8.5/3.3 ms; flip angle = 12; matrix = 512 × 512 × 384; voxel size = 0.47 × 0.47 × 0.5). For the fMRI, a gradient-echo T2*-weighted echo-planar imaging sequence was used in the acquisition of 150 functional volumes on the verb generation task (TR/TE = 2500/30 ms; flip angle = 70; matrix = 64 × 64 × 30; voxel size = 3.75 × 3.75 × 4), and a different sequence was used in the acquisition of 374 functional volumes during the stop-signal task and 210 functional volumes in the resting-state (TR/TE = 2000/30 ms; flip angle = 70; matrix = 64 × 64 × 27; voxel size = 3.75 × 3.75 × 4.5).

Image processing

Task-based fMRI processing

Task functional images were processed using the Statistical Parametric Mapping software package (SPM12; Wellcome Trust Centre for Neuroimaging, London, UK). Preprocessing followed the default pipeline and included: (a) alignment of each participant’s fMRI data to the AC-PC plane by using the anatomical image; (b) head motion correction, where the functional images were realigned and resliced to fit the mean functional image; (c) co-registration of the anatomical image to the mean functional image; (d) re-segmentation of the transformed anatomical image using a symmetric tissue probability map; (e) spatial normalization of the functional images to the MNI (Montreal Neurological Institute, Montreal, Canada) space with 3 mm3 resolution; and (f) spatial smoothing (FWHM = 4 mm).

The general linear model for the verb generation task was defined for each participant by contrasting activation control blocks. The general linear model for the stop-signal task was defined for each participant by contrasting correct ‘stop’ correct ‘go’ trials. For both tasks, the BOLD (Blood-Oxygen-Level-Dependent) signal was estimated by convolving the task’s block/trial onsets with the canonical hemodynamic response function. Six motion realignment parameters extracted from head motion preprocessing were included as covariates of no interest, and a high-pass filter (128 s) was applied to the contrast images to eliminate low-frequency components.

After that, we utilized the VOI analysis function within SPM12 to extract the first eigenvariate of the BOLD time courses from the stop-signal contrast images, which highly resembles the mean time course across the voxels. This extraction was carried out region-wise for the main components of the inhibitory control network: pars opercularis, pars triangularis, preSMA, and STN. Pars opercularis and pars triangularis of the IFC were defined following the criteria of the Harvard-Oxford atlas44–47. For the PreSMA region, we used the SMA region, as defined by the same atlas, but including only voxels anterior to MNI Y = 048. STN was defined by a 10mm box centered at MNI coordinates 10, −15, −522,49.

Lastly, we computed whole-brain voxel-wise functional asymmetry maps from the stop-signal contrast images. To do so, stop-signal contrast images were flipped at midline, inverting the right and left hemispheres, and subsequently subtracted from the original unflipped contrast images50.

Resting-state fMRI processing

Resting-state functional images were processed using the Data Processing Assistant for the Resting-State toolbox (DPARSFA)51, which is based on SPM and the Data Processing & Analysis of Brain Imaging toolbox (DPABI)52. Preprocessing steps included: (a) slice-timing correction for interleaved acquisitions; (b) head motion correction (no participant had a head motion of more than 2 mm maximum displacement in any direction or 2° of any angular motion throughout the scan); (c) co-registration of the anatomical image with the mean functional image; (d) new segmentation to DARTEL; (e) removal of nuisance variance through linear regression: six parameters from the head motion correction, white matter signal, cerebrospinal fluid signal, and global mean signal; (f) spatial normalization to the MNI (3 mm3); (g) spatial smoothing (FWHM = 4 mm); (h) removal of the linear trend in the time series; (i) band-pass temporal filtering (0.01–0.1); (j) normalization to a symmetric template; (k) Voxel-Mirrored Homotopic Computation (VMHC)53, calculated as the Pearson correlation coefficient of every voxel with its hemispheric counterpart; and (l) normalization of all voxel-wise time courses to Fisher z values. Finally, mean VMHC values were extracted for the pars opercularis, pars triangularis, preSMA, and STN. ROI definition can be found in the previous section.

Structural MRI processing

Structural images were processed via voxel-based morphometry (VBM) using the CAT12 toolbox54, which is based on SPM. Preprocessing steps followed the recommended pipeline and included: (a) segmentation into grey matter, white matter, and cerebrospinal fluid; (b) registration to the ICBM standard template; (c) modulated normalization of grey matter and white matter segments to the MNI template; (d) spatial smoothing (FWHM = 6 mm); and (e) extraction of Regions of Interest (ROIs) values from native space. Three ROIs were delimited by the voxels mapped as genu, body and splenium of the corpus callosum according to the Mori atlas55. Total intracranial volume was also extracted for use as covariate of no interest.

Individual assessment of functional lateralization

Individual functional lateralization was assessed by calculating the Laterality Index (LI) on the unflipped contrast images34,35. We used the bootstrap method implemented in the LI-toolbox56, based on SPM. LI is a proportion of the brain activation between the two hemispheres, thus giving us information about the direction and degree of hemispheric specialization during a particular function in a single individual. LI ranges from +1 (totally leftward function) to −1 (totally rightward function). For both the verb generation task and the stop-signal task, we explored the LI of the inferior frontal region roughly corresponding to the classic Broca’s area: pars opercularis and pars triangularis of the inferior frontal gyrus, according to the Harvard-Oxford atlas44–47. This region is critical for both language production3 and response inhibition24, depending on the hemisphere. During the verb generation task, language production was classified as typically lateralized (LI higher than +0.4) or atypically lateralized (LI lower than +0.4). We used +0.4 as a cut-off point (contrary to the more traditional +0.2), based on previous findings that emphasized the importance of lateralization strength when grouping individuals6,30.

Statistical analyses

Task-based region-wise analysis

We performed different analyses to investigate the hypothesis that the atypical group would show a mirrored brain organization during the stop-signal task. First, we conducted a region-wise analysis to compare the functional asymmetry of the primary components within the inhibitory control network between the typically and atypically lateralized groups. To achieve this, we employed a repeated-measures MANOVA, with Hemisphere (left/right) and Region (pars opercularis/pars triangularis/preSMA/STN) as within-subject factors and Group (typical/atypical) as the between-subject factor. Subsequently, to assess potential hemispheric differences in any of these regions between both groups, we contrasted the simple effects of the Hemisphere × Group interaction across the Region factor. All resulting p-values were adjusted using a Bonferroni correction.

Task-based voxel-wise analyses

Previous analysis was complemented with whole-brain functional asymmetry contrast maps (extracted from the stop-signal task) that were compared in typically and atypically lateralized groups. To do so, we performed a whole-brain two-sample t-test (voxel-wise P < 0.001; FWE cluster-corrected at P < 0.05) via SPM12.

We also explored the functional overlap between both tasks in the subgroup of participants that presented an ambilateral organization of at least one function (n = 48). To do so, we overlayed two one-sample t-tests restricted to the IFC: one for the stop-signal task (voxel-wise P < 0.001; FWE cluster-corrected at P < 0.05), and one for the verb generation task (voxel-wise P < 0.05; uncorrected). Functional overlap was then expressed as the proportion of overlapped voxels for every function and hemisphere.

Correlational analyses

The relationship between the hemispheric specialization of the verb generation task and the stop-signal task at the IFC was also examined linearly. More precisely, we conducted a Pearson’s correlation between the LI from the verb generation task and the LI from the stop-signal task used in the individual lateralization classification (P < 0.05, two-tailed).

Using these LIs, we also tested if a linear relationship existed between the functional lateralization of an individual, and several behavioral tests and neuroanatomical measures. Partial Spearman’s correlations were computed between the LI of both tasks and: stop-signal task performance (RT, accuracy, and SSRT), SPQ derived-scores, AQ derived-scores, reading test performance (length and familiarity effects on speed and accuracy), functional connectivity measures (VMHC), and white matter volume of the callosal genu. Age and general intelligence were included as covariates in all analyses. Total intracranial volume was also included as covariate in the volumetric analysis.

Results

Inhibitory control shifts its lateralization according to language production

Inhibitory control components during the stop-signal task were defined using four Regions Of Interest (ROIs) and subsequently analyzed through a MANOVA. This analysis included Hemisphere and Region as within-subject factors and Group (typical/atypical) as a between-subject factor. Examination of the simple effects across each level of Region revealed that all the tested structures displayed a statistically significant Hemisphere × Group interaction (pars opercularis F1,84 = 25.59, P < 0.001; pars triangularis F1,84 = 19.95, P < 0.001; preSMA F1,84 = 7.63, P = 0.028; STN F1,84 = 14.24, P < 0.001; P values Bonferroni-adjusted; see Fig 1a). In other words, depending on the lateralization group, all four regions appeared to exhibit a leftward (atypical group) or rightward (typical group) lateralization of their BOLD signal during the ‘correct stop > correct go’ condition of the stop-signal task.

Hemispheric lateralization of inhibitory control according to language lateralization.

(A) ROI analysis of the main components of the inhibitory control network. Graphs depict adjusted mean BOLD signal during successful inhibitions on the stop-signal task (‘stop > go’ contrast) for both hemispheres and both groups. All four structures showed significant Hemisphere × Group interactions. Error bars represent one standard error. pOper = pars opercularis, pTri = pars triangularis, preSMA = presupplementary motor area, and STN = subthalamic nucleus. (B) Voxel-wise whole-brain analysis of functional asymmetry. Significance maps (voxel-wise P < 0.001; FWE cluster-corrected at P < 0.05; color bar represents t value) are displayed in three-dimensional reconstructions plus coronal and transversal slices using MNI space. (C) Mean BOLD values of the significant regions found in the voxel-wise whole-brain analysis. Graphic depicts BOLD values for every region, hemisphere, and group. IFC = inferior frontal cortex, preSMA = presupplementary motor area, AngG = angular gyrus, MTG = middle temporal gyrus, Thal = thalamus, STN = subthalamic nucleus, Caud = caudate. (D) Functional overlap between language production and inhibitory control in the IFC of ambilateral participants. Overlapping maps for inhibition (voxel-wise P < 0.001; FWE cluster-corrected at P < 0.05) and language (voxel-wise P < 0.05; uncorrected) are displayed in coronal, sagittal and transversal slices using MNI space. VGT = verb generation task, SST = stop-signal task.

A voxel-wise whole-brain analysis of the functional asymmetry maps during the stop-signal task was performed to complement the ROI results. Two-sample t-test comparisons of the typical and atypical groups during the ‘correct stop > correct go’ condition showed a differential asymmetry pattern in all the ROI-corresponding areas, while also revealing differences in other cortical and subcortical regions, including the angular gyrus, middle temporal gyrus, and caudate (Fig 1b). Specifically, the typical group presented a rightward activity in these regions, whereas the atypical group exhibited a leftward shift (Fig 1c).

Finally, the Pearson’s correlation between the frontal LIs (pars opercularis + pars triangularis) from the verb generation task and the stop-signal task revealed a strong inverse correlation (r84 = −.58; two-tailed P < 0.001; R2 = .34) (Fig 2). Out of 50 typically lateralized for language, 33 (66%) presented a typical right lateralization for inhibitory control, and 17 (34%) showed an atypical (left – ambilateral) lateralization. Among the 36 individuals from the atypical group, only 8 (22.2%) presented a typical lateralization during the stop-signal task, whereas 28 (77.8%) showed an atypical lateralization. Importantly, no participant presented a leftward + leftward or rightward + rightward segregation pattern for language production and response inhibition. Strong segregation of both functions into different hemispheres was observed in 38 participants (44.2%). The remaining 48 individuals (55.8%) presented an ambilateral representation of one or both functions, with an important level of spatial overlap in their activations (see Fig 1d). In this ambilateral group, language production presented an overlap in 32.3% of its voxels in the left hemisphere, and in 58.6% of its voxels in the right hemisphere. Inhibitory control showed an overlap in 57.2% of its voxels in the left hemisphere, and in 22.8% of its voxels of the right hemisphere.

Correlation between the LIs of the verb generation and stop-signal tasks.

r = −.583, two-tailed P < 0.001, R2 = .339. Negative values indicate rightward lateralization, whereas positive values indicate leftward lateralization. Segregated and ambilateral phenotypes are also depicted according to the background color. The green area corresponds to segregated individuals (both functions strongly lateralized), and the red area corresponds to integrated individuals (at least one function ambilaterally controlled). Numbers inside each quadrant denote the number of individuals (n) contained in it. Each individual datapoint is symbolized according to its functional organization: 〇 = typical segregation; Ill = reversed segregation; △ = ambilateral inhibition; Ill = ambilateral language; Ill = ambilateral language and inhibition.

Laterality Indexes correlate to interhemispheric functional connectivity, callosal volume and preclinical markers

Correlational analyses revealed certain links between the functional lateralization of both tasks and behavioral, neuroanatomical and connectivity measures (Table 1).

First, we studied how the inter-hemispheric functional connectivity and callosal size behaved in relation to LI. Volume in the genu, body and splenium of the corpus callosum was inversely related with language LI (respectively: ρ83 = −.25, P = 0.02; ρ83 = −.26, P = 0.02; and ρ83 = −.25, P = 0.02). However, only the callosal genu was found related with inhibition LI (ρ83 = .25, P = 0.02). That is, the volume of the callosal genu increased as functional organization of the IFC became more atypical, extending this effect to the whole corpus callosum when considering exclusively language lateralization. On the same line, ROI interhemispheric functional connectivity analyses at rest revealed that VMHC of the pars triangularis also increased as a function of inhibition LI (ρ77 = .26, P = 0.02), but a similar relation failed to be found for language LI (ρ77 = −.12, P = 0.3). No statistically significant linear relationships were found when exploring the VMHC of the pars opercularis, the preSMA, or the STN. It should be mentioned, however, that the association between VMHC of the STN and inhibition LI was close to significance (ρ77 = .22, P = 0.054).

Behaviorally, performance (RT, SSRT and accuracy) during the scanner stop-signal task did not correlate with language and inhibition LIs. Regarding the reading test, accuracy for word length and word familiarity were found correlated with both LIs (length and language: ρ84 = −.24, P = 0.03; length and inhibition: ρ84 = .21, P = 0.04; familiarity and language: ρ84 = −.23, P = 0.03; familiarity and inhibition: ρ84 = .3, P = 0.01). That is, an atypical organization was related with a higher rate of errors when reading long words and unfamiliar words. No significant correlations were found when exploring reading speed. SPQ score was found significantly related with inhibition LI (ρ82 = .25, P = 0.02), but it did not reach statistical significance when paired with language LI (ρ84 = −.19, P = 0.08). AQ, on the other hand, presented an association with language LI (ρ84 = −.25, P = 0.02) but not with inhibition LI (ρ84 = .08, P = 0.5).

Discussion

The present study employed the stop-signal task to investigate brain areas involved in the inhibitory control function in two different groups of individuals: typically and atypically lateralized for language production. As expected, the atypical participants showed, as a group, a mirrored brain organization of the inhibitory control function compared to the typical group. This leftward organization affected the entire inhibitory control network, including the IFC, preSMA, and STN. However, some participants manifested a clear overlap in the control of language and inhibitory functions. Our results also demonstrate that atypical organization of language production is associated with an increased white matter volume of the corpus callosum, and that atypical lateralization of inhibitory control is related with a higher interhemispheric functional coupling of the IFC. Behaviorally, atypical lateralization was not associated with performance on the task, but an association existed with worse reading accuracy, and more schizotypy and autistic traits.

Our first relevant result is that the group of individuals with atypical lateralization of language presented a mirrored brain organization during the stop-signal task compared to the typically lateralized group. Remarkably, the hemispheric asymmetry differences obtained affected the entire inhibitory control brain network25. The inferior frontal cortex, or IFC, associated with the initiation of the stop command and the main cortical area of this system, showed a clear lateralization effect, with the typical group almost exclusively using the right part and the atypical group mostly activating the left part. The presupplementary motor area, or pre-SMA, related to the implementation of the stop command, was more bilaterally involved in both groups, although atypicals activated the left part more and typicals activated the right part more. Importantly, the hemispheric reversal of organization also affected the subcortical structures participating in this network, namely the STN and thalamus. These structures displayed a lateralization pattern similar to the preSMA, with a more pronounced bilateral engagement but higher involvement of the right part in typicals and of the left part in atypicals. Thus, this is the first demonstration that an entire cerebral network controlling a cognitive function shows a completely flipped organization in an atypical population for language, which supports the Causal hypothesis of lateralization10,11. Moreover, this result reveals that hemispheric lateralization goes beyond cerebral cortices, and even interhemispheric connectivity, through the corpus callosum, given that it also involves frontal-subcortical circuits. Additionally, the specificity of this result to the areas of the proposed network is another element that supports the relevance of this system in inhibitory control.

It should be noted that previous studies have demonstrated a significantly higher involvement of the left IFC during cognitive control tasks in bilingual individuals when compared to monolinguals57,58. However, this could not be directly considered in the current study due to the limited number of monolinguals participants (n = 8), which would not allow for statistically robust tests. So, we cannot rule out the possibility that bilingualism could be acting as an enabling mechanism among atypically lateralized participants for the shift of the inhibitory control network to the left hemisphere. Therefore, the findings of this study should be extrapolated to the monolingual population with caution.

Remarkably, this pattern of reversed specialization for inhibitory control was observed in the absence of a relation between functional lateralization and performance differences during the ‘go’ or ‘stop’ conditions. Even though the ‘go’ condition involves a lexical decision requiring language processing, and the ‘stop’ condition consists of inhibiting a linguistic process, the functional organization of the IFC during the tasks did not affect the response speed (RT and SSRT) or the ‘go’ accuracy in either case. Thus, brain organization for language and inhibition, as initially defined in our study, and cognitive efficiency are not directly related, based on our data. This finding is consistent with prior studies highlighting that: (1) a densely segregated functional organization, rather than a reversed organization, is linked to general17 and specific26 cognitive deficits; and (2) specific cognitive deficits appear to be confined to the domains of verbal and spatial skills26.

Importantly, despite observing a complimentary relationship between language production and inhibitory control at the group level, not all the individuals presented a strong and segregated distribution of these two functions. The vast majority of hemispherically segregated individuals showed the typical segregation of language in the left hemisphere and inhibitory control in the right hemisphere, whereas only a low percentage displayed the opposite segregated organization. The rest, in agreement with similar reports on other functions19, lacked a clear hemispheric lateralization of language production, inhibitory control, or both. In all cases, this implied some level of overlapping or sharing of the same area of the inferior frontal cortex. Both the low ratio of atypical segregation and the high ratio of bilateral inhibitory control could be explained by the fact that typical right hemispheric functions (such as inhibition) seem to be more bilaterally represented than typical left functions (such as language)59. Some conclusions can be extracted from this: (1) Although infrequent, a strong atypical segregation of the two functions is possible; (2) Our data have not revealed any case with a segregation of both frontal cognitive functions in the same hemisphere; and (3) A large percentage of participants with a strong left lateralization of language presented a bilateral control of inhibition, showing that a typical dominance for language does not imply a strongly lateralized organization of other cognitive functions. The existence of exceptions to the group pattern of mirrored brain organization requires us to use caution when interpreting these data as supporting the causal hypothesis. Although the negative correlation between the lateralization indexes during the verb generation task and the stop-signal task is strong at the group level, some individual data support the statistical hypothesis12 instead. Given that we found no cases of both cognitive functions completely sharing a hemisphere (see Fig 2), we propose that causal-supporting vs. statistical-supporting results reflects two different pathways to cognitive control. On the one hand, the segregated pathway (in line with the Causal hypothesis) develops due to the cerebral bias towards lateralizing certain cognitive functions in a single hemisphere. On the other hand, the integrated pathway (in line with the Statistical hypothesis), appears when ontogenetic development towards strong lateralized control fails in some way.

These findings could have significant implications when studying neonatal lesions that facilitate the development of left-handedness60, or when assessing the consequences of brain lesions in the left-handed population61. For example, a left-sided lesion in an individual with rightward language lateralization may result in inhibitory control and social deficits62. It also raises questions about neuroplasticity, particularly how functional reorganization following a brain lesion (such as a left-to-right reorganization of language) would impact this hemispheric relationship between language and inhibitory control, and the cognitive consequences for inhibitory control after such a reorganization63. These results could also offer valuable insights for cognitive rehabilitation procedures in patients with psychiatric disorders that seem to be linked to a higher prevalence of left-handedness and alterations in language lateralization64.

The corpus callosum is the main cerebral structure in interhemispheric connectivity, with the genu area being responsible for the connectivity between the two inferior frontal gyri65. Phylogenetic data have demonstrated a negative association between the corpus callosum volume and hemispheric specialization, suggesting that the ontogenetic development of the brain is supported by a decrease in interhemispheric connectivity to potentiate and establish intrahemispheric connectivity and hemispheric specialization66. There is evidence of this mechanism in language development67 and in agenesis of the corpus callosum, which has been associated with lower lateralization of language and worse language performance68,69. In line with this proposal, we have replicated the relation between callosal enlargement and atypical functional organization found in a previous study30. Moreover, in our data, this structural correlation was accompanied by an association between atypical lateralization of inhibitory control and strength of interhemispheric functional connectivity in the pars triangularis of the IFC.

As hypothesized, when considering the functional lateralization during both tasks, atypical organizations were found linearly related with higher scores on a schizotypy test (only for inhibitory control), higher scores on an autistic spectrum test (only for language production), and a higher rate of reading errors (both language production and inhibitory control). The first result is consistent with previous structural, functional, and behavioral data that have shown reduced language lateralization in schizophrenia patients8,70. In the case of schizotypy, the evidence for reduced lateralization of cognitive functions is mixed and may respond to differences in the methodology employed71. However, consistently with the present results, this condition has been previously associated to impaired behavioral and neural processing during the stop-signal task72, and to an abnormal frontal functional asymmetry73. The second result relates atypical language lateralization with the presence of autistic traits, a result consistent with data obtained in autism74. The third result shows that atypical organization was related with a higher rate of reading errors when considering word length and word familiarity. The length effect has been considered a pathognomonic symptom of reading disorders such as developmental dyslexia75 and pure alexia76. These disorders have also been characterized by a weak language lateralization7. Thus, presented results support the hypothesis that atypical hemispheric specialization is related with worse cognitive performance in the linguistic domain, and even preclinical traits of some neurodevelopmental disorders among healthy population.

In conclusion, our study demonstrates a strong connection between the lateralization of language production in the IFC and the hemispheric specialization of the inhibitory control network. This functional interrelationship is apparent through both regional and voxel-wise analyses and impacts all the primary cerebral hubs of inhibitory control. Furthermore, the atypical lateralization of these two cognitive functions shows a linear correlation with increased inter-hemispheric functional connectivity in the IFC, a larger corpus callosum, the presence of subclinical markers of schizotypy and autism, and decreased reading accuracy. Consequently, our research provides compelling evidence in support of the causal and corpus callosum hypotheses of lateralization among left-handers, and offers valuable insights into the pathogenic mechanisms underlying atypical lateralization in healthy population.